---
title: "Week 3 Notes Confidence Interval"
author: "Shu Guo"
date: "Wednesday, February 11, 2015"
output: html_document
---

## Introduction
We do not recommend reporting p-values as the only statistical summary of your results. The reason is simple: statistical significance does not guarantee scientific significance. With large enough sample sizes, one might detect a statistically significance difference in weight of, say, 1 microgram. But is this an important finding? Would we say a diet results in higher weight if the increase is less than a fraction of a percent? The problem with reporting only p-values is that you will not provide a very important piece of information: the effect sizes.

An much more attractive alternative is to report confidence intervals. A confidence interval includes relays information about your estimated effect size and the uncertainty associated with this estimate. Here we use the mice data to illustrate the concept behind confidence intervals.

## Confidence interval for population mean

We show how to construct a confidence interval for the population mean of control female mice. We start by reading in the data and selecting the appropriate rows:
```{r}
dat <- read.csv("mice_pheno.csv")
chowPopulation <- dat[dat$Sex=="F" & dat$Diet=="chow",3]
```
The population average $\mu_X$ is our parameter of interest here:

```{r}
mu_chow <- mean(chowPopulation)
print(mu_chow)
```
We are interested in estimating this parameter. In practice we do not get to see the entire population so, as we did for p-values, we demonstrate how we can use samples to do this. Let's start with a sample of size 30:
```{r}
N <- 30
hf <- sample(chowPopulation,N)
```
We know this is a random variable, so the sample average will not be a perfect estimate. In fact, because in this illustrative example and we know the value of the parameter, we can see that they are not exactly the same. A confidence interval is a statistical way of reporting our finding, the sample average, in a way that explicitly summarizes the variability of our random variable.

With a sample size of 30, we will use the CLT. The CLT tells us that $\bar{X}$ or `mean(hf)` follows a normal distribution with mean $\mu_X$ or `mean(chowPopulation)` and standard error approximately $s_X/\sqrt{N}$ or

```{r}
se <- sd(hf)/sqrt(N)
print(se)
```

<a name="interval"></a>

